---
title: "Bryan Rincon: Breaking Down the Breakout"
subtitle: "Swing changes, offensive growth, and the case for an everyday shortstop"
author: "Andy Lucas"
format:
  html:
    embed-resources: true
execute:
  echo: false
  message: false
  warning: false
---

## Does the Bat Make Him an Everyday Shortstop?

After watching Rincon live against Altoona on June 23 and on video
throughout 2026, I see the actions, range, and arm versatility to stay
at shortstop. His fielding percentage dropped as his workload increased,
so consistency deserves another look, but the tools are there.

The bigger question is the bat. Rincon showed more power, 
a lower strikeout rate, and more walks in Double A. Do those gains make him a future
everyday shortstop? I’ll look at his production and swing changes to
assess what looks sustainable and what still needs to hold up against
better pitching.

## Recommendation and Confidence

I would raise Rincon’s offensive projection toward an everyday shortstop outcome, with moderate confidence in the bat. The strongest evidence is increased game power alongside lower strikeout frequency and sustained walk production.

```{r}
library(baseballr)
library(dplyr)
library(knitr)
library(kableExtra)
library(plotly)
```

```{r}
rincon_2025 <- readRDS("data/raw/rincon_2025.rds")
rincon_2026 <- readRDS("data/raw/rincon_2026.rds")
```

```{r}
# Combine the seasons and calculate rates from the original counts
rincon_comparison <- bind_rows(rincon_2025, rincon_2026) |>
  mutate(k_pct = 100 * strike_outs / plate_appearances,
    bb_pct = 100 * base_on_balls / plate_appearances,
    hr_pct = 100 * home_runs / plate_appearances,
    iso = (total_bases - hits) / at_bats,
    babip_calc = (hits - home_runs) /
      (at_bats - strike_outs - home_runs + sac_flies)) |>
  select(
    season, sport_abbreviation, plate_appearances,
    k_pct, bb_pct, hr_pct, iso, babip_calc, ops)

rincon_comparison |>
  as.data.frame() |>
  knitr::kable(
    format = "html",
    caption = "Bryan Rincon: 2025–2026 Offensive Performance",
    col.names = c(
      "Season", "Level", "PA", "K%", "BB%",
      "HR/PA (%)", "ISO", "BABIP", "OPS"),
    digits = c(0, 0, 0, 1, 1, 1, 3, 3, 3),
    row.names = FALSE,
    align = c("l", "l", rep("r", 7))) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center",
    font_size = 16) |>
  kableExtra::row_spec(
    0,
    bold = TRUE,
    color = "white",
    background = "#002D62") |>
  kableExtra::row_spec(
    2,
    bold = TRUE,
    color = "#B52832",
    background = "#FFF1F2")
```

Rincon showed more game power while lowering his strikout rate and drawing
more walks. His strikeout rate fell from 28.7% to 20.8%, his walk rate
climbed from 13.9% to 17.8%, and his ISO increased from .117 to .250.
The combination is more encouraging than the power jump alone: he
produced substantially more damage without an increase in strikeout
frequency. His BABIP also rose from .250 to .299, so improved results on
balls in play contributed to the breakout. The move from High A to
Double A and the change in playing environment remain important context.

## Did the Improvements Hold Throughout 2026?

To determine whether the breakout extended beyond a hot stretch, I
examined his strikeout, walk, and home run rates by month. Each month is
compared with his 2025 season rates, with plate appearances shown to
give the results proper sample context.

```{r}
# Load the saved game logs
game_log_response <- jsonlite::fromJSON(
  "data/raw/rincon_2026_game_logs.json",
  flatten = TRUE)

rincon_games <- game_log_response$stats$splits[[1]]

# Label each game by its date
monthly_comparison <- rincon_games |>
  mutate(
    date = as.Date(date),
    month = format(date, "%Y-%m")) |>
  group_by(month) |>
  summarise(
    PA = sum(stat.plateAppearances),
    AB = sum(stat.atBats),
    H = sum(stat.hits),
    TB = sum(stat.totalBases),
    SO = sum(stat.strikeOuts),
    BB = sum(stat.baseOnBalls),
    HR = sum(stat.homeRuns),
    .groups = "drop") |>
  mutate(
    k_pct = 100 * SO / PA,
    bb_pct = 100 * BB / PA,
    hr_pct = 100 * HR / PA,
    iso = (TB - H) / AB) |>
  arrange(month)
```
Navy lines show his 2026 monthly rates. Red dashed lines show
his 2025 season rates. Hover over a dot for the percentage
and plate appearances.
```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

# Put the monthly rates into plotting format
monthly_chart_data <- monthly_comparison |>
  mutate(
    month_label = paste0(
      format(as.Date(paste0(month, "-01")), "%b"),
      "\n", PA, " PA"),
    month_label = factor(month_label, levels = month_label)) |>
  tidyr::pivot_longer(
    cols = c(k_pct, bb_pct, hr_pct),
    names_to = "metric",
    values_to = "rate")
# Use the 2025 season rates as reference lines
reference_2025 <- rincon_comparison |>
  as.data.frame() |>
  filter(season == "2025") |>
  select(k_pct, bb_pct, hr_pct) |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "rate")
metric_order <- c("k_pct", "bb_pct", "hr_pct")

monthly_chart_data <- monthly_chart_data |>
  mutate(metric = factor(metric, levels = metric_order))

reference_2025 <- reference_2025 |>
  mutate(metric = factor(metric, levels = metric_order))

monthly_plot <- ggplot2::ggplot(
  monthly_chart_data,
  ggplot2::aes(x = month_label, y = rate, group = 1)) +
  ggplot2::geom_hline(
    data = reference_2025,
    ggplot2::aes(yintercept = rate),
    color = "#B52832",
    linetype = "dashed") +
  ggplot2::geom_line(color = "#002D62", linewidth = 1) +
  ggplot2::geom_point(
    ggplot2::aes(
      text = paste0(
        "Month: ", month,
        "\nPA: ", PA,
        "\nRate: ", sprintf("%.1f%%", rate))),
    color = "#002D62",
    size = 2.5) +
  ggplot2::facet_wrap(
    ~metric,
    nrow = 1,
    labeller = ggplot2::as_labeller(c(
      k_pct = "Strikeout rate",
      bb_pct = "Walk rate",
      hr_pct = "Home run rate"))) +
  ggplot2::scale_y_continuous(
    limits = c(0, 35),
    labels = function(x) paste0(x, "%")) +
  ggplot2::labs(
    title = "Rincon’s Offensive Performance by Month",
    subtitle = "Navy: 2026 Double-A monthly rates | Red dashed: 2025 High-A season rates",
    x = NULL,
    y = "Percentage of plate appearances",
    caption = "Source: MLB Stats API. Different levels and environments limit comparisons.") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold"))
plotly::ggplotly(monthly_plot, tooltip = "text")
```

Rincon’s lower strikeout frequency and higher home run frequency
persisted across every month of 2026 relative to his 2025 season rates.
His walk rate increased each month after May. May paired a comparatively
lower walk rate (11.8%) with a higher home run rate (7.1%), raising the
possibility of a more aggressive approach; however, swing rates, count
data, or video would be needed to test that explanation. Overall, the
improvement was not confined to one hot month, although smaller monthly
samples, particularly September’s 49 plate appearances, and different
playing environments limit the comparison.

## Mechanical Look: Bat Setup and Timing

### Changes from 2024 through 2026

In the 2024 footage I reviewed, Rincon started with a flatter barrel
near his shoulder. From the left side, the bat appeared to rest on the
shoulder; from the right side, it hovered above it. He then brought the
barrel into a more upright position during his load.

By 2025, his starting bat position looked much closer to the 2026 setup.
The difference in the selected clips was more subtle: I observed more
hand movement during his 2025 load, while the 2026 move appeared
quieter.

The progression looks like a gradual refinement rather than a complete
overhaul in one offseason. The larger change in barrel orientation
appears between the 2024 footage and the later clips, with further
reduction in hand movement visible in the selected 2026 swings. Matching
the stage of the delivery and reviewing additional plate appearances
would help confirm how consistently those differences appear.

### Reported Work with Kevin Long

The article attributed to NBC Sports Philadelphia adds a specific
developmental adjustment to the video observations. Rincon described
getting too handsy and allowing his swing to get long when trying to
correct it. During spring training in 2026, Kevin Long introduced a
drill using the cue of taking the knob toward the baseball to simplify
his hand path.

This gives context to the work on his swing, but the article does not
identify the more upright starting barrel as an instruction from Long.
Rincon also credited getting stronger and gaining confidence. His return
from the hamate injury provides additional context, although the
available evidence cannot separate the effects of health, strength, and
mechanical changes.

### Potential Effect on the Swing

My read is that the offseason between 2025 and 2026 gave Rincon time to
settle into the more upright barrel position already visible in his 2025
setup. As he recovered, gained strength, and worked on his hand path, he
may have become more comfortable with that position and reduced the
extra hand movement during his load. The selected 2026 swings appear
quieter into launch, which could help him repeat his timing and deliver
the barrel more consistently. This is my interpretation of the video
progression, supported by the reported strength and hand path work,
rather than a confirmed explanation of how the adjustment developed.

The statistical results support taking that possibility seriously. He
produced more damage while striking out less often. However, the lower
strikeout rate does not establish improved contact on swings, and the
selected footage cannot confirm a shorter swing or better plate
coverage. Those remain priorities for the next video review.

### What I Would Look for Next

The next look would focus on whether the setup produces a more
repeatable move into launch. I would compare his hands and barrel at the
same checkpoints in the delivery, particularly pitcher release and front
foot plant, using multiple plate appearances from each side.

I would include swings against velocity and breaking balls, along with
takes, fouls, and misses. The key question is whether the adjustment
helps him get the barrel to different locations and adjust to changes in
speed. The selected highlights suggest a setup change, but they do not
yet establish improved plate coverage or a shorter swing.

## Updated Projection

I would raise Rincon’s offensive projection and give greater weight to
an everyday major league shortstop outcome, with moderate confidence in
the bat. The combination of more game power, fewer strikeouts per plate
appearance, and a higher walk rate makes this a meaningful step forward.
Those gains also extended beyond one hot stretch.

My live look against Altoona on June 23, 2026, combined with video
review throughout the season, gives me confidence that he can stay at
shortstop. He shows smooth actions approaching the ball, covers ground,
and makes throws from different arm slots and body positions. His speed
also contributes on the bases. That foundation gives him a path to an
everyday role without requiring the bat to carry his entire profile.

The decline in fielding percentage deserves attention alongside
his increased workload, but the defensive tools still look
intact. More innings do not explain the rate decline on their
own. I would review the errors for patterns in routine execution
and throwing accuracy. For now, I view consistency as an area
to monitor rather than a reason to move him off shortstop.

The remaining uncertainty is how much offense will translate against
better pitching. His quieter load and reported work on the hand path
offer a plausible explanation for the gains, but the next test is
whether he maintains his timing against velocity and adjusts to breaking
balls from both sides of the plate. I see a stronger case for an
everyday shortstop than I did before the breakout, while leaving the
offensive ceiling open until he faces more advanced competition.

## Limitations

Access to 2024 and 2025 footage was very limited. Those clips provide
reference points, but they do not establish how consistently Rincon used
each setup or exactly when adjustments occurred. Differences in camera
angle and the stage of the delivery also affect comparisons of his hands
and barrel.

Most of the available 2026 clips reviewed were highlights of home runs
or other extra base hits. This creates a sample weighted toward his best
swings and outcomes. It is useful for examining how his swing looks when
he does damage, but cannot establish how often he repeats those
movements, handles difficult pitches, or makes contact. A fuller review
would include takes, fouls, misses, and unsuccessful plate appearances
from both sides.

The statistical comparison spans different levels and playing
environments without adjustments for league or park effects. Monthly
samples are smaller than season samples, particularly September’s 49
plate appearances. Lower strikeout frequency does not by itself
establish improved contact rate or pitch recognition.

The connections drawn between his setup, hand path work, physical
development, and offensive results are scouting interpretations rather
than demonstrated causes. Similarly, fielding percentage does not
measure his full defensive value or establish why his error rate
changed. The role projection combines the available statistics,
reporting, and my live and video observations, with further evaluation
needed against more advanced competition.

<details>

<summary>Sources and Scouting Looks</summary>

#### Statistical Sources

- **MLB Stats API via baseballr:** 2025 High A and 2026 Double A regular season batting totals. The 2026 game logs were downloaded directly from the API. Retrieved October 1, 2026. Original inputs are preserved in `data/raw`.

- **Calculations in R:** Strikeout rate, walk rate, home run rate, ISO, BABIP, and monthly comparisons were calculated from the underlying counts. Game log totals were checked against the 2026 season totals.

- **[MiLB: Bryan Rincon career fielding statistics](https://www.milb.com/player/bryan-rincon-804858?stats=career-r-fielding-minors&year=2026):** Shortstop appearances, innings, errors, and fielding percentage used to contextualize his defensive workload and results. Fielding percentage is reported directly by MiLB. Accessed October 1, 2026.

#### Video Sources

- **[MLB Film Room: Bryan Rincon](https://www.mlb.com/video/?q=bryan+rincon&qt=FREETEXT):** Selected footage used to examine his setup, barrel orientation, and load across seasons and from both sides of the plate. Accessed October 1, 2026.

- **[Jersey Shore BlueClaws and Bryan Rincon Instagram post](https://www.instagram.com/p/DJ5U0aFOBj6/?img_index=2):** Posted May 20, 2025. The second carousel item was used to examine his setup and load from the left side. Accessed October 1, 2026.

#### Reporting

- **[MLB: Phillies prospect Bryan Rincon’s breakout season at Double A](https://www.mlb.com/amp/news/phillies-prospect-bryan-rincon-breakout-season-double-a.html):** Background on his offensive development and prior injuries.

- **[Rising shortstop prospect soaks up CBP, earns Phillies’ top minors honor](https://sports.yahoo.com/articles/rising-shortstop-prospect-soaks-cbp-172000939.html):** NBC Sports Philadelphia article republished by Yahoo Sports. Used for Rincon’s comments about working with Kevin Long, the hand path drill, strength gains, and confidence.

#### Personal Scouting Looks

- Live look against Altoona on June 23, 2026, supplemented by video review throughout the season. Observations of his defensive actions, throwing versatility, and speed, along with the projection that he can remain at shortstop, are my own evaluations.

</details>
